Supplementary Information

### 2021-08-10 doc/manuscript/response_to_reviews.Rmd

I have a few minor remarks that the authors can consider:

Lines 178-180 please insert citations for this statement. I suggest also consider artefact of sampling biases caused by the absence of old, fast-growing trees (cf. “slow-grower survivorship bias”) and of young, slow-growing trees (cf. “big-tree selection bias”) in the dataset ( see Duchesne, L., Houle, D., Ouimet, R., Caldwell, L., Gloor, M., & Brienen, R. (2019). Large apparent growth increases in boreal forests inferred from tree-rings are an artefact of sampling biases. Scientific reports, 9(1), 1-9.).

**Because the relevant references for this statement overlap significantly with those for the previous sentence, we have merged those two sentences: "Yet, even after correcting for ontogeny, growth trend detection remains subject to various potential sampling and analysis biases, which are fundamentally driven by the limitation that it is not possible to use a contemporary set of tree cores to obtain a representative sample of a species's population at all time points throughout the history of a dynamically changing stand [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbass-ahles_influence_2014; @sullivan_effect_2016; @bowman_detecting_2013; @brienen_tree_2017; @hember_tree_2019; @duchesne_large_2019]." Note that we have added the Duchense reference here.**

**We have also added mention of slow-grower survivorship bias in the discussion: "For instance, older trees, which provide the only records available for the earliest decades, may be competitive winners that had above-average growth rates within their cohorts [@aubry-kientz_joint_2015], which would upwardly bias average growth rate estimates for early decades ["juvenile selection effect", @groenendijk_no_2015]. In contrast, the oldest age classes being dominated by trees with below-average growth rates [e.g., @piovesan_lessons_2019] could downwardly bias average growth rate estimates for early decades ["a slow-grower survivorship bias", @brienen_detecting_2012; @duchesne_large_2019]."**

@@

Lines 180-182 please quote at least one supporting reference.

**We have slightly modified this sentence and added a reference: "Tree growth rates are sensitive to stand dynamics, with competition -- the intensity of which tends to increase as forests mature -- reducing woody growth rates [e.g., @zhang_halfcentury_2015]."**

Lines 646 -648 in temperate closed old-growth forests an opposite behaviour is often reported for the oldest trees. Slow but increasing long-term growth allows for maximum longevity not only in beech, but also in oaks and pines   (see Piovesan, G., Biondi, F., Baliva, M., De Vivo, G., Marchiano, V., Schettino, A., & Di Filippo, A. (2019). Lessons from the wild: slow but increasing long‐term growth allows for maximum longevity in European beech. Ecology, 100(9), e02737 and references therein).

**We have modified this section to provide an example of the opposite behaviour, citing the reference mentioned (and @duchesne_large_2019, mentioned above): "For instance, older trees, which provide the only records available for the earliest decades, may be competitive winners that had above-average growth rates within their cohorts [@aubry-kientz_joint_2015], which would upwardly bias average growth rate estimates for early decades ["juvenile selection effect", @groenendijk_no_2015]. In contrast, the oldest age classes being dominated by trees with below-average growth rates [e.g., @piovesan_lessons_2019] could downwardly bias average growth rate estimates for early decades ["a slow-grower survivorship bias", @brienen_detecting_2012; @duchesne_large_2019]."**

### 2021-08-10 doc/manuscript/ForestGEO_treerings.Rmd
finish responses to R1

Our analytical framework reconstructs growth changes in a sampled tree population over time while accounting for climate, DBH, and persistent growth differences among individuals (Fig. `r schematic`), thereby addressing some important challenges to obtaining unbiased estimates of growth trends attributable to non-climatic environmental drivers.
First, we account for changes in climate that may drive directional growth trends.
For example, dramatic growth declines at LT, driven by a strong regional warming and drying trend [@touchan_millennial_2011; @williams_temperature_2013], are in part factored out by accounting for the primary climate drivers, such that a significant decline over time was detected for only one of the two species (Fig. `r year_fig`).
Second, we show that growth rate -- by any metric -- varies nonlinearly with DBH and with patterns dependent upon the species and environmental context (Fig. `r size`), reinforcing the concept that growth trend analyses should incorporate cross-sectional analyses to correct for growth ontogeny [e.g., through regional curve standardization, @peters_detecting_2015].
Our method does this, differing from the conceptually parallel method of regional curve standardization in that we standardize relative to DBH rather than age, correct for any trends in the most influential climate drivers, and include random effects of tree to account for persistent growth differences among individuals.
The latter addresses a third important challenge, as those growth differences among individuals can bias estimated growth trends in positive or negative directions [@brienen_detecting_2012;@nehrbass-ahles_influence_2014; @groenendijk_no_2015; @brienen_tree_2017; @vandersleen_trends_2017].
For instance, older trees, which provide the only records available for the earliest decades, are competitive winners that may have had above-average growth rates [@aubry-kientz_joint_2015], which would upwardly bias average growth rate estimates for early decades [@groenendijk_no_2015].
By including a random effect of tree, our approach likely reduces the most severe potential biases associated with persistent growth differences across individuals [@brienen_detecting_2012; @brienen_tree_2017; @bowman_detecting_2013], yet observed trends nevertheless represent only the sampled population of trees, as opposed to tree populations at all points throughout the time frame analyzed.
Within this context, signals of changing growth rate over time are attributable to some combination of stand dynamics (e.g., recruitment and succession, changing stand structure) and environmental drivers (e.g., indirect effects of climate change, rising atmospheric CO~2~, deposition of SO~2~ and NO~x~).

>>>>

Our analytical framework reconstructs growth changes in a sampled tree population over time while accounting for climate, DBH, and persistent growth differences among individuals (Fig. `r schematic`), thereby addressing some important challenges to obtaining unbiased estimates of growth trends attributable to non-climatic environmental drivers.
First, we account for changes in climate that may drive directional growth trends.
For example, dramatic growth declines at LT, driven by a strong regional warming and drying trend [@touchan_millennial_2011; @williams_temperature_2013], are in part factored out by accounting for the primary climate drivers, such that a significant decline over time was detected for only one of the two species (Fig. `r year_fig`).
Second, we show that growth rate -- by any metric -- varies nonlinearly with DBH and with patterns dependent upon the species and environmental context (Fig. `r size`), reinforcing the concept that growth trend analyses should incorporate cross-sectional analyses to correct for growth ontogeny [e.g., through regional curve standardization, @peters_detecting_2015].
Our method does this, differing from the conceptually parallel method of regional curve standardization in that we standardize relative to DBH rather than age, correct for any trends in the most influential climate drivers, and include random effects of tree to account for persistent growth differences among individuals.
The latter addresses a third important challenge, as those growth differences among individuals can bias estimated growth trends in positive or negative directions [@brienen_detecting_2012;@nehrbass-ahles_influence_2014; @groenendijk_no_2015; @brienen_tree_2017; @vandersleen_trends_2017].
For instance, older trees, which provide the only records available for the earliest decades, may be competitive winners that had above-average growth rates within their cohorts [@aubry-kientz_joint_2015], which would upwardly bias average growth rate estimates for early decades ["juvenile selection effect", @groenendijk_no_2015]. 
In contrast, the oldest age classes being dominated by trees with below-average growth rates [e.g., @piovesan_lessons_2019] could downwardly bias average growth rate estimates for early decades ["a slow-grower survivorship bias", @brienen_detecting_2012; @duchesne_large_2019].
By including a random effect of tree, our approach likely reduces the most severe potential biases associated with persistent growth differences across individuals [@brienen_detecting_2012; @brienen_tree_2017; @bowman_detecting_2013], yet observed trends nevertheless represent only the sampled population of trees, as opposed to tree populations at all points throughout the time frame analyzed.
Within this context, signals of changing growth rate over time are attributable to some combination of stand dynamics (e.g., recruitment and succession, changing stand structure) and environmental drivers (e.g., indirect effects of climate change, rising atmospheric CO~2~, deposition of SO~2~ and NO~x~).
